Transfer learning for echelle spectroscopy

### 2021-11-10 paper/paper1/ms.tex
writing prose about the method, continued!

We start with a high resolution pre-computed stellar model spectrum, $\mathsf{M}(\bm{\lambda})$.  The procedure is largely agnostic to the exact details of how this spectrum was made.  For the purposes of this paper, we will showcase examples from two well-known families of precomputed synthetic spectra: \texttt{PHOENIX} \citep{husser13} for stellar spectra $(T_{\mathrm{eff}}\in [2300, 10000]\;K)$ and \texttt{Sonora} \citep{2021ApJ...920...85M} for brown dwarf spectra $(T_{\mathrm{eff}}\in [1000, 2300]\;K)$.  We place the following demands on the the precomputed spectra.  They should have sporadic regions of discernable continuum devoid of lines, and the continuum must vary smoothly in wavelength.  The spectral lines or pseudo-lines should be resolved, and not sub-sampled.  We suspect most stellar spectra meet these criteria, with the exception of the coolest M dwarfs and brown dwarfs.  The method can hypothetically handle spectra without discernable continua, but we suspect some modifications to the preprocessing steps would be needed.  We truncate the red and blue limits of the precomputed synthetic spectrum to match a high-bandwidth echelle spectrograph, extended with a buffer at the edges of size $\pm \Delta \lambda_{\mathrm{buffer}}$, chosen to account for plausible radial velocity and rotational broadening of real stars.  A generous buffer of $v \sin{i} < 500 \;\kmps$ and $|RV|<500 \;\kmps$ yields a typical buffer of about 30~\AA.

The choice of limiting the bandwidth to a region-of-interest around a single echelle spectrograph bandwidth stems from computational constraints.  In principle there is no fundamental limit to the bandwidth one could clone with the method presented here, up to and including the entire precomputed synthetic spectral model bandwidth.  We adopt the exact native wavelength sampling with no smoothing or interpolation, yielding a wavelength vector $\bm{\lambda}$ with length $N_\mathrm{pixels}$ equal to the number of pixels within the extents of our region of interest including the buffers.

@@

We initialize the cloned model with a series of preprocessing steps.  We divide the entire spectrum by a black body of the same effective temperature $T_{\mathrm{eff}}$ as the model template.  The resulting signal typically possesses a smooth continuum variation.  An additional continuum flattening step ensures that subsequent spectral line finding steps get applied uniformly.  Here, we identify the continuum regions as the most prominent peaks in the specturm separated by a pixel distance of at least $15000-90000$ pixels from adjacent peaks, yielding between 60 and 5 points.  We then fit an $n_{\mathrm{poly}} = 3^{rd}-5^{th}$ order polynomial to these peaks, evaluate the polynomial on the entire bandwidth, and divide the spectrum by this trend.  Visual inspection of this continuum-flattened spectrum typically possesses variations at the $<0.2\%$ level.  Tuning the pixel distance and polynomial order can lower the undulations to $<0.05\%$ level for some input spectra.

We emphasize that a recreation of the unvarnished input spectrum can be obtained by multiplying the continuum-flattened signal by the ``perturbed black body'', $\mathsf{B}(\bm{\lambda})$, that symbolizes the black body modulated by continuum opacity or broad-band radiative transfer effects.  This smooth spectrum may be useful for applications that need to keep track of broad-band flux, such as low resolution spectrographs, or regions with molecular bandheads.  The ``perturbed black body'' continuum model $\mathsf{B}$ contains  $n_{\mathrm{poly}}+1$ fixed lookup parameters, plus the input $T_{\mathrm{eff}}$.

We apply a local-minimum-finding algorithm on the spectrum by defining a prominence threshold $P_{rom} \in (0.005, 0.1)$.  This threshold dictates the number of lines that will be modeled: a lower prominence finds more, smaller lines, and a larger prominence finds fewer, deeper lines.  The prominence algorithm successfully finds lines that reside on top of broad line wings, or unresolved bandheads provided that the individual lines exceed the prominence threshold in their local region.  The number of lines $N_{\mathrm{lines}}$ depends on the bandwidth, prominence, and the intrinsic properties of the input spectrum, principally effective temperature and metallicity.

For this paper, we illustrate examples for two \'echelle spectrographs with particularly large spectral grasp: the Habitable Zone Planet Finder \citep[HPF,][]{2014SPIE.9147E..1GM} on the Hobby-Eberly Telescope at McDonald Observatory in Fort Davis, Texas; and the Immersion Grating Infrared Spectrograph \citep[IGRINS,][]{park14} currently on the Gemini South Telescope on Cerro Pach\'on in Chile.  The $R=55,000$ HPF has a native bandwidth of $x-y$, that we expand to $x-y^*$ including the edge buffers.  IGRINS has two cameras for $H$ and $K$ band, with the combined spectrum spanning $z-y$ including the edge buffers and the region in-between the two cameras, all at a resolving power of $(R=45,000)$.  The spectrograph acquisition, reduction, and post-processing steps yield data $\mathsf{D}(\bm{\lambda}_{d})$, where $\bm{\lambda}_{d}$ is the wavelength vector at the instrumental resolution and sampling of each instrument, generally much coarser than the resolution and sampling grid of the precomputed synthetic spectra.  The data wavelength vector also contains pesky gaps between \'echelle orders, whereas the precomputed grid is usually contiguous.

Figure XX shows how the number of detected lines $N_{\mathrm{lines}}$ scales with effective temperature and prominence threshold for the \texttt{PHOENIX} grid, truncated to the bandwidths-plus-buffers for HPF and IGRINS.

\begin{outline}
  \1 Define a prominence threshold and find locations of peaks above that threshold
  \2 Derive coarse properties about those peaks: widths and amplitudes
  \1 We now have a coarse model that needs to be fine-tuned with autodiff
\end{outline}

The low-pass filter should be set to capture the genuine spectral shape, without over-fitting broad line-wings such as those in deep Hydrogen and sodium lines.  We have experimentally found it helpful to hand-tune models of exceptionally broad and non-standard line-wings with with an optional pre-processing step.

>>>>

We start with a high resolution pre-computed synthetic stellar model spectrum, $\mathsf{S}(\bm{\lambda})$.  The procedure is largely agnostic to the exact details of how this spectrum was made.  For the purposes of this paper, we will showcase examples from two well-known families of precomputed synthetic spectra: \texttt{PHOENIX} \citep{husser13} for stellar spectra $(T_{\mathrm{eff}}\in [2300, 10000]\;K)$ and \texttt{Sonora} \citep{2021ApJ...920...85M} for brown dwarf spectra $(T_{\mathrm{eff}}\in [1000, 2300]\;K)$.  We place the following demands on the the precomputed spectra.  They should have sporadic regions of discernable continuum devoid of lines, and the continuum must vary smoothly in wavelength.  The spectral lines or pseudo-lines should be resolved, and not sub-sampled.  We suspect most stellar spectra meet these criteria, with the exception of the coolest M dwarfs and brown dwarfs.  The method can hypothetically handle spectra without discernable continua, but we suspect some modifications to the preprocessing steps would be needed.  We truncate the red and blue limits of the precomputed synthetic spectrum to match a high-bandwidth echelle spectrograph, extended with a buffer at the edges of size $\pm \Delta \lambda_{\mathrm{buffer}}$, chosen to account for plausible radial velocity and rotational broadening of real stars.  A generous buffer of $v \sin{i} < 500 \;\kmps$ and $|RV|<500 \;\kmps$ yields a typical buffer of about 30~\AA.

The choice of limiting the bandwidth to a region-of-interest around a single echelle spectrograph bandwidth stems from computational constraints.  In principle there is no fundamental limit to the bandwidth one could clone with the method presented here, up to and including the entire precomputed synthetic spectral model bandwidth.  We adopt the exact native wavelength sampling with no smoothing or interpolation, yielding a wavelength vector $\bm{\lambda}_s$ with length $N_\mathrm{s}$ equal to the number of pixels within the extents of our region of interest including the buffers.

@@

We initialize the cloned model with a series of preprocessing steps.  We divide the entire spectrum by a black body of the same effective temperature $T_{\mathrm{eff}}$ as the model template.  The resulting signal typically possesses a smooth continuum variation.  An additional continuum flattening step ensures that subsequent spectral line finding steps get applied uniformly.  Here, we identify the continuum regions as the most prominent peaks in the specturm separated by a pixel distance of at least $15000-90000$ pixels from adjacent peaks, yielding between 60 and 5 points.  We then fit an $n_{\mathrm{poly}} = 3^{rd}-5^{th}$ order polynomial to these peaks, evaluate the polynomial on the entire bandwidth, and divide the spectrum by this trend.  Visual inspection of this continuum-flattened spectrum typically possesses variations at the $<0.2\%$ level.  Tuning the pixel distance and polynomial order can lower the undulations to $<0.05\%$ level for some input spectra.  The high-pass filtering step should be set to capture the genuine spectral shape, without over-fitting broad line-wings such as those in deep Hydrogen and sodium lines.  We have experimentally found it helpful to hand-tune models of exceptionally broad and non-standard line-wings with an optional pre-processing step, not described here.

We emphasize that a recreation of the unvarnished input spectrum can be obtained by multiplying the continuum-flattened signal by the ``perturbed black body'', $\mathsf{B}(\bm{\lambda}_s)$, that symbolizes the black body modulated by continuum opacity or broad-band radiative transfer effects.  This smooth spectrum may be useful for applications that need to keep track of broad-band flux, such as low resolution spectrographs, or regions with molecular bandheads.  The ``perturbed black body'' continuum model $\mathsf{B}$ contains  $n_{\mathrm{poly}}+1$ fixed lookup parameters, plus the input $T_{\mathrm{eff}}$.

Next we identify the spectral lines.  We apply a local-minimum-finding algorithm on the spectrum by defining a prominence threshold $P_{rom} \in (0.005, 0.1)$.  This threshold dictates the number of lines that will be modeled: a lower prominence finds more, weaker lines, and a larger prominence finds fewer, deeper lines.  The prominence algorithm successfully finds lines that reside on top of broad line wings, or unresolved bandheads provided that the individual lines exceed the prominence threshold in their local region.  The number of lines $N_{\mathrm{lines}}$ depends on the bandwidth, prominence, and the intrinsic properties of the input spectrum, principally effective temperature and metallicity.

For this paper, we illustrate examples for two \'echelle spectrographs with particularly large spectral grasp: the Habitable Zone Planet Finder \citep[HPF,][]{2014SPIE.9147E..1GM} on the Hobby-Eberly Telescope at McDonald Observatory in Fort Davis, Texas; and the Immersion Grating Infrared Spectrograph \citep[IGRINS,][]{park14}
currently on the Gemini South Telescope on Cerro Pach\'on in Chile.  The $R=55,000$ HPF has a native bandwidth of $x-y$, that we expand to $x-y^*$ including the edge buffers.  IGRINS has two cameras for $H$ and $K$ band, with the combined spectrum spanning $z-y$ including the edge buffers and the region in-between the two cameras, all at a resolving power of $(R=45,000)$.  The spectrograph acquisition, reduction, and post-processing steps yield data $\mathsf{D}(\bm{\lambda}_{d})$, where $\bm{\lambda}_{d}$ is the wavelength vector at the instrumental resolution and sampling of each instrument, generally much coarser than the resolution and sampling grid of the precomputed synthetic spectra.  The data wavelength vector may also contains gaps between \'echelle orders, whereas the precomputed grid is usually contiguous.  HPF may have up to $2048\times28=57,344$ pixels, and IGRINS has typically about $75,000$ pixels, after common truncation of noisy edge pixels.

Figure XX shows how the number of detected lines $N_{\mathrm{lines}}$ scales with effective temperature and prominence threshold $P_{rom}$ for the \texttt{PHOENIX} grid, truncated to the bandwidths-plus-buffers for HPF and IGRINS.  The HPF-truncated spectra have $N_s=XX$ and the IGRINS-truncated spectra have $N_s=XX$.

So far we have only one piece of information about the peaks: their location.  Next, we derive coarse properties about each detected peak: its amplitude and width, again using the prominence algorithms implemented in \texttt{scipy} \citep{2020SciPy-NMeth}.

There does not exist a general-purpose, single-shot algorithm for obtaining the lineshape in the presence of overlapping spectral lines: where do the wings of one line begin and the wings of another adjacent line end?  We therefore do not attempt to determine anything about the lineshape at this stage, and instead assume that the lines resemble a Voigt profile, with a guess width about equally split between Lorentzian and Gaussian.

We have now arrived at the \emph{blas\'e} clone model $\mathsf{M}(\bf{\lambda}_s)$ for a flattened synthetic spectrum $\mathsf{S}$: it is the cumulative product of transmission through the sea of all overlapping spectral lines:

\begin{eqnarray}
  \mathsf{M}(\bm{\lambda}_s) = {\displaystyle \prod_{j=1}^{N_{\mathrm{lines}}}  (1-a_j V(\bm{\lambda}_s-\lambda_{c,j}, \sigma_j, \gamma_j) )}
\end{eqnarray}

where $V(\lambda, \sigma_j, \gamma_j)$ is the normalized Voigt profile with Gaussian standard deviation $\sigma$, Lorentzian half-width $\gamma$, at line center position $\lambda_c$, for the $j^{th}$ spectral line.  The amplitude $a$ is equal to the area under the curve of the line.

The model evaluated with its coarse initial values would have terrible performance: it would only vaguely resemble the synthetic spectral model, with up to $\pm 50\%$ undulations from the inexact assignment of widths,lineshapes, and amplitudes.  Instead, we tune the parameters of the model, starting from these coarse initial values.  This model has $N_{\mathrm{lines}}\times 3$ free parameters, where the center wavelength is held fixed and the amplitude, width, and lineshape are allowed to vary.

As seen in Figure XX, the number of lines can exceed $XX$ depending on the
